Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-01-04 LaTeX/Turnover.tex
nothing

% Including pdf figures
\usepackage{graphicx}
\graphicspath{ {../Plots/} }
\usepackage{pdfpages}
%really place a figure in a location
\usepackage{float}
%Overrun caption
\usepackage[CaptionAfterwards]{fltpage}
% Math stuff
\usepackage{amsmath}
% Bibliographies
\usepackage[numbers]{natbib}
\bibpunct{(}{)}{,}{a}{}{;}

@@

\newpage
\linenumbers
\modulolinenumbers[2]

\begin{abstract}
Sex determination systems are remarkably dynamic; many studied taxa display transitions of sex-determining genes between chromosomes or the evolution of new sex-determining systems. 
%Predominant theories in which new sex-determining systems are favoured by selection involve sex ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection). 
Here, we utilize population genetic models to study the spread of novel sex-determining systems in the presence of
%where we also include 
%a period of selection during the haploid stage produced by one sex, e.g., pollen or sperm competition. 
haploid gametic selection, e.g., pollen or sperm competition.
Haploid selected loci experience a form of sex-specific selection (because gametic competition occurs predominantly among haploids produced by males) and can cause sex ratios at birth to become biased (because sex ratios are determined by the fertilization success of X- versus Y-bearing pollen/sperm). 
%We find that the evolution of sex determination systems where mothers determine sex at birth (e.g., environmental sex determination where sex is determined at birth) is influenced by classic Fisherian sex ratio selection. {\color{red}(Maybe not true???)}
%However, notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
Notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
In addition, we find that linkage of an ancestral sex chromosome to a locus under haploid selection can favour transitions between male and female heterogamety (e.g., XY to ZW), which is not the case for any forms of diploid sex specific selection (e.g., sexually antagonistic selection).
During these transitions, new sex-determining alleles spread despite breaking up favourable associations that build up between ancestral sex-determining loci and selected loci, reducing population mean fitness. 
Furthermore, a period of selection among haploids can favour the stable maintenance of polymorphic sex determination systems. 
Thus, our models offer several new insights to be explored as information about sex determination in non-model taxa accumulates.

@@

Animals and angiosperms exhibit extremely diverse sex determination systems \citep[reviewed in][]{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric/dioecious clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex determination systems can be found in the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.
In addition, multiple transitions have occurred between genetic and environmental sex determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

Here, we use mathematical models to find the conditions under which new sex determination systems are favoured by selection when there is a period of selection among haploid gametes/gametophytes. 
Selection among haploid genotypes is thought to occur primarily among pollen/sperm, which can compete whenever there are more pollen/sperm than required for fertilization \citep{Mulcahy:1996ha,JOSEPH:2004haa}. 
Haploid selection may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.
\textcolor{red}{add something about meiotic drive here?}

@@

%genetic system
We consider a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a region under selection, with alleles $A$ and $a$.
And locus \textbf{M} is a novel sex-determining region, with alleles $M$ and $m$.
With genotype $MM$ at locus \textbf{M}, the sex of a zygote is determined by the genotype at locus \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males).
With at least one $m$ allele at locus \textbf{M}, a zygote develops as a female with probability $k$ and as a male with probability $1-k$.
With $k=0$, the novel sex determiner is a masculinizer (e.g., a neo-Y chromosome) and with $k=1$ the novel sex determiner is a feminizer (e.g, a neo-W chromosome).
With intermediate $k$, locus \textbf{M} is interpreted as an offspring controlled environmental sex-determining region.
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
The life-cycle begins with competition between haploid gametes/gametophytes (hereafter gametes) from each sex, where selection depends on the sex of the diploid they came from and their allele at the \textbf{A} locus.
Gametes from males then pair randomly with gametes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (or the \textbf{M} genotype of their mother in the case of maternal control).
Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
This is followed by meiosis with recombination and sex-specific meiotic drive. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We model meiotic drive at the viability locus only; individuals of sex $i$ who are heterozygous at the \textbf{A} locus produce haploids bearing allele $A$ with probability $\alpha^i$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

%what were going to do, briefly
With allele $M$ fixed at the \textbf{M} locus, sex is determined by locus \textbf{X} and an equilibrium is reached at locus \textbf{A}. %where, importantly, locus \textbf{A} can be stably polymorphic. %the dynamics can be described by the frequency of $A$ in gametes/gametophytes from the homogametic sex and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex.
We then examine the ability of a rare, novel sex-determiner, $m$, to increase in frequency from this equilibrium.
Numerical simulations are used to examine when an invading mutation goes to fixation.

@@

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes from the homogametic sex (e.g., in eggs from an XX female, $p^f_X$), the frequencies of $A$ in the two types of gametes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$), and the frequencies of the sex-determining factors in gametes from the heterogametic sex (e.g., the frequency of Y in sperm, $q$).
Assuming selection and meiotic drive are weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete are small, as is the bias in the sex-determining factor from the heterogametic sex, and we can solve for the mean frequency of $A$ across all types ($p_A$), the difference in the frequencies of $A$ between two of the three types, and the bias in the frequency of the sex-determining factor, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
Without haploid selection or meiotic drive our results reduce to those of \cite{vanDoorn:2007eu}. %when $k=0$ (neo-$Y$ invading a $ZW$ system) or when $k=1$ (neo-$W$ invading an $XY$ system) (\textcolor{red}{CHECK}).
However, with haploid selection or meiotic drive a stable polymorphism at locus \textbf{A} no longer requires sexually antagonistic selection. %, as it can also be achieved by ploidally-antagonistic selection.

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming no competition among gametes from females, no meiotic drive in females, and linear arrangement \textbf{X}\textbf{A}\textbf{M}. %(i.e., $\chi = r R$). %(SAY SOMETHING HERE ABOUT OTHER ARRANGEMENTS?). %equal dominance coefficients in the two sexes,
Equivalent results for invasions into a $ZW$ system can be derived by consistently switching the roles of males and females \citep{vanDoorn:2010hu}.

%\subsubsection*{Neo-$Y$}
%
%%write in general form without weak selection
%A rare, dominant neo-$Y$ ($k=0$) is always expected to invade the ancestral $XY$ system when the average growth rate of mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
%When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the neo-$Y$ does not invade.
%Otherwise, the mutant allele increases in frequency on one \textbf{A} background and declines on the other and neo-$Y$ invasion requires
%\begin{equation}
%R \left[ \frac{p^f_X w_{m,a}}{g_A^*} + \frac{(1 - p^f_X) w_{m,A}}{g_a^*} \right] w_{m,Aa} < \nu_m,
%\end{equation}
%where $w_{m,i}$ is the relative viability fitness of males depending on their haploid or diploid genotype at the \textbf{A} locus (with $i\in\{A,a,AA,Aa,aa\}$) and
%$\nu_m = 
%  p^f_X p_{Ym} w^m_{a} w^m_{AA} +
%  p^f_X (1-p_{Ym}) w^m_{a} w^m_{AA} +
%  (1-p^f_X) p_{Ym} w^m_{a} w^m_{AA} +
%  (1-p^f_X) (1-p_{Ym}) w^m_a w^m_{AA}$
%is the mean relative fitness of resident males.
%%and $g_j^*$ is the growth rate of $X^m$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).
%Neo-$Y$ invasion therefore occurs for any recombination rate, $R$, when the net flow of double recombinants is from the less fit to the more fit \textbf{A} background (making the term inside the square brackets negative).
%%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^f_X) w_{m,A} g_a^* < p^f_X w_{m,a} g_A^*$.
%When the net flow of double recombinants is from the more fit to the less fit haplotype, neo-$Y$ invasion can still occur when the rate of recombination is small enough.
%
%%weak selection
%Assuming weak selection, we can solve for the invasion fitness of the neo-$Y$ explicitly, giving
%\begin{equation}
%\lambda_{Y,XY} \approx 1 + V_A S_A (r - R)s^f,
%\end{equation}
%where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $S_A = s^f/ \left[ r R \left( s^f + s^m \right)^2 t_m^2 \right]$ (INTERPRETATION), $s^f$ and $s^m$ are the respective selection coefficients for $A$ in diploid females and males, $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males, and we've assumed equal dominance coefficients in the two sexes.
%The neo-$Y$ can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s^f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%
%\subsubsection*{Neo-$W$}
%
%%write in general form without weak selection
%Similarly, a rare, dominant neo-$W$ ($k=1$) will invade the ancestral $XY$ system whenever the average growth rate of $XAm$ and $Xam$ haplotypes is positive, $(g_A + g_a) / 2 > 1$. 
%When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the neo-$W$ does not invade.
%Otherwise neo-$W$ invasion requires
%\begin{equation}
%R \left[ \frac{\bar{p}_m w_{m,A}}{g_a^*} + \frac{(1 - \bar{p}_m) w_{m,a}}{g_A^*} \right] w_{f,Aa} < \nu_f
%\end{equation}
%where $\bar{p}_m = (p_{Ym}+p^m_X) / 2$ is the mean frequency of $A$ in gametes/gametophytes from males, $w_{f,i}$ is the relative viability fitness of females depending on their diploid genotype at the \textbf{A} locus (with $i\in\{AA,Aa,aa\}$), and 
%$\nu_f = 
%  p^f_X p^m_X w^m_{a} w_{AA,f} +
%  p^f_X (1-p^m_X) w^m_{a} w_{Aa,f} +
%  (1-p^f_X) p^m_X w^m_{a} w_{Aa,f} +
%  (1-p^f_X) (1-p^m_X) w^m_a w_{aa,f}$
%is the mean relative fitness of resident females. 
%%The rate of recombination between the \textbf{X} and \textbf{A} loci, $r$, does not explicitly appear because females are homozygous at the \textbf{X} locus with an ancestral $XY$ system.
%As in the case of the neo-$Y$, neo-$W$ invasion therefore occurs with any recombination rate, $R$, when the net flow of recombinants is from the less fit to the more fit haplotype.
%%Assuming $g_A^*>0$ and $g_a^*<0$ without loss of generality, this is $(1 - \bar{p}_m) w_{m,a} g_a^*< \bar{p}_m w_{m,A} g_A^*$.
%And when the net flow of recombinants is from the more fit to the less fit  haplotype, neo-$W$ invasion can still occur when the rate of recombination is small enough.
%
%%weak selection
%Assuming weak selection, we can solve for the invasion fitness of the neo-$W$ explicitly, giving
%\begin{equation}
%\lambda_{W,XY} \approx 1 + V_A S_A \frac{(2r(1-R)-R) s^f + (1-2r)R s^m }{2},
%\end{equation}
%where we've once again assumed equal dominance coefficients in the two sexes.
%In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than is the ancestral sex-determining locus, $r<R$, a novel sex-determiner can invade.
%For example, with $R=1/2$ the neo-$W$ invades if there is any linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t_m\neq0$), and there is selection for or against $A$ in both males and females ($s^m s^f >0$) that is stronger in females than in males ($|s^f| > |s^m|$).

\subsubsection*{Neo-$Y$ and neo-$W$}

%write in general form without weak selection
A rare, dominant neo-$Y$ ($k=0$) or neo-$W$ ($k=1$) is always expected to invade an ancestral $XY$ system when the average growth rate of the mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the new sex-determining allele does not invade.

Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ \frac{p^f_X w^m_a (1-\alpha^m)}{g_A^*} + \frac{(1 - p^f_X) w^m_A \alpha^m}{g_a^*} \right] w^m_{Aa} < \nu^m,
\end{equation}
for the neo-$Y$, and 
\begin{equation}\label{eq:neoWR}
R \left[  \frac{(1 - \bar{p}^m) w^m_{a}}{g_A^*} + \frac{\bar{p}^m w^m_{A}}{g_a^*} \right] w^f_{Aa} < \nu^f
\end{equation}
for the neo-$W$.
Here $\bar{p}^m = (p^m_X+p^m_Y)/2$ is the average frequency of $A$ in gametes produced by males, $w^i_{j}$ is the relative viability fitness of sex $i\in\{m,f\}$ depending on their haploid or diploid genotype at the \textbf{A} locus (with $j\in\{A,a,AA,Aa,aa\}$) and
$\nu^i = 
  p^f_X p^m_{\textbf{X}^i} w^m_{a} w^i_{AA} +
  p^f_X (1-p^m_{\textbf{X}^i}) w^m_{a} w^i_{Aa} +
  (1-p^f_X) p^m_{\textbf{X}^i} w^m_{a} w^i_{Aa} +
  (1-p^f_X) (1-p^m_{\textbf{X}^i}) w^m_a w^i_{aa}$,
with $\textbf{X}^m = Y$ and $\textbf{X}^f = X$, is the mean relative fitness of resident individuals of sex $i$.
%and $g_j^*$ is the growth rate of $X^m$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).
Although male meiotic drive does not explicitly appear in Equation \ref{eq:neoWR}, it does affect the average frequency of $A$ in gametes from males, $\bar{p}^m$, and thus can play a role in neo-W invasion.

Equations \eqref{eq:neoYR} and \eqref{eq:neoWR} show that the new sex-determining allele is expected to invade for any recombination rate, $R$, when the net flow of double recombinants is from the less fit to the more fit \textbf{A} background (making the terms inside the square brackets in Equations \ref{eq:neoYR} and \ref{eq:neoWR} negative).
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^f_X) w_{m,A} g_a^* < p^f_X w_{m,a} g_A^*$.
When the net flow of double recombinants is from the more fit to the less fit haplotype, the new sex-determining allele can still invade when the rate of recombination between it and the selected locus, $R$, is small enough.

%weak selection
Assuming weak selection and meiotic drive, we can explicitly solve for the invasion fitness of the new sex-determining allele, $m$, into the ancestral $XY$ system, giving
\begin{equation}
\lambda_{m,XY} \approx 1 + V_A S_A C_m,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus.
We will consider haploid selection and meiotic drive separately.
With haploid selection and no drive we have $S_A = s^f (t^m)^2 / \left[ r R \left( s^f + s^m \right)^2 \right]$, where $s^f$ and $s^m$ are the respective selection coefficients for $A$ in diploid females and males, $t^m$ is the selection coefficient for $A$ in gametes from males, and we've assumed equal dominance coefficients in the two sexes.
With drive and no haploid selection we replace $t^m$ with $\alpha^m-1/2$.

For the neo-$Y$ we have $C_Y = (r - R)s^f $ with haploid selection and $C_Y = 4(r - R)s^f $ with drive.
The neo-$Y$ can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$), there is diploid selection in females ($s^f\neq0$), and there is either haploid selection between gametes from males ($t^m\neq0$) or meiotic drive in males ($\alpha^m\neq1/2$).
This reduces to the weak-linkage results of \cite{vanDoorn:2007eu} when we do not assume equal dominance coefficients in the two sexes, there is no haploid selection or meiotic drive, and there is free recombination between locus \textbf{A} and one of the sex-determining regions.

For the neo-$W$ we have $C_W = [(2r(1-R)-R) s^f + (1-2r)R s^m]/2$ with haploid selection and $C_W = 4[r(1-2R) s^f + (1-2r)R s^m]/2$ with drive.
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than the ancestral sex-determining locus is, $r<R$, a neo-$W$ can invade (Figure \ref{fig:Winvasion}).
For example, $R=1/2$ the neo-$W$ invades if there is any linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes in males ($t^m\neq0$), and there is selection for or against $A$ in both males and females ($s^m s^f >0$) that is stronger in males than in females ($|s^m| > |s^f|$).
With meiotic drive and $R=1/2$, all that is required for neo-W invasion is $r<1/2$ and $s^m s^f >0$.
Our results reduce to the weak-linkage results of \cite[][Equation 3]{vanDoorn:2010hu} when we do not assume equal dominance coefficients in the two sexes and there is no haploid selection or meiotic drive.

>>>>

% Including pdf figures
\usepackage{graphicx}
\usepackage{pdfpages}
%really place a figure in a location
\usepackage{float}
%Overrun caption
\usepackage[CaptionAfterwards]{fltpage}
% Math stuff
\usepackage{amsmath}
% Bibliographies
\usepackage[numbers]{natbib}
\bibpunct{(}{)}{,}{a}{}{;}

@@

\begin{abstract}
Sex determination systems are remarkably dynamic; many studied taxa display transitions of sex-determining genes between chromosomes or the evolution of new sex-determining systems. 
%Predominant theories in which new sex-determining systems are favoured by selection involve sex ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection). 
Here, we utilize population genetic models to study the spread of novel sex-determiners in systems with 
%a period of selection during the haploid stage produced by one sex, e.g., pollen or sperm competition. 
haploid gametic selection, e.g., pollen or sperm competition.
Haploid selected loci experience a form of sex-specific selection (because gametic competition occurs predominantly among haploids produced by males) and can also cause sex ratios at birth to become biased (because sex ratios are determined by the fertilization success of X- versus Y-bearing pollen/sperm). 
We find that the evolution of sex determination systems where mothers determine sex at birth (e.g., environmental sex determination where sex is determined at birth) is influenced by classic Fisherian sex ratio selection. {\color{red}(Maybe not true???)}
However, notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
In addition, we find that linkage of an ancestral sex chromosome to a locus under haploid selection can favour transitions between male and female heterogamety (e.g., XY to ZW), which is not the case for any forms of diploid sex specific selection (e.g., sexually antagonistic selection).
During these transitions, new sex-determining alleles spread despite breaking up favourable associations that build up between ancestral sex-determining loci and selected loci, reducing population mean fitness. 
Furthermore, a period of selection among haploids can favour the stable maintenance of polymorphic sex determination systems. 
Thus, our models offer several new insights to be explored as information about sex determination in non-model taxa accumulates.

@@

Animals and angiosperms exhibit extremely diverse sex determination systems, as reviewed by \citet{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric/dioecious clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex determination systems can be found the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.
In addition, multiple transitions have occurred between genetic and environmental sex determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

Here, we use mathematical models to find the conditions under which new sex determination systems are favoured by selection where we include a period of selection among haploid gametes/gametophytes. 
Selection among haploid genotypes is thought to occur primarily among pollen/sperm, which can compete whenever there are more pollen/sperm than required for fertilization \citep{Mulcahy:1996ha,JOSEPH:2004haa}. 
Haploid selection may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
\textcolor{red}{add something about drive}

@@

%genetic system
We will consider a three locus model.
Locus \textbf{X} is the ancestral sex-determining region (SDR), with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a region under selection, with alleles $A$ and $a$.
And locus \textbf{M} is a novel sex-determining region, with alleles $M$ and $m$.
With genotype $MM$ at locus \textbf{M}, the sex of a zygote is determined by the genotype at locus \textbf{X} ($XX$ become females and $XY$ males, or $ZW$ become females and $ZZ$ males).
With at least one $m$ allele at locus \textbf{M}, a zygote develops as a female with probability $k$ and as a male with probability $1-k$.
With $k=0$, the novel sex determiner is a masculinizer (e.g., a neo-Y chromosome) and with $k=1$ the novel sex determiner is a feminizer (e.g, a neo-W chromosome).
With intermediate $k$, locus \textbf{M} is interpreted as an environmental sex-determining region.

%life-cycle
The life-cycle begins with competition between haploid gametes/gametophytes within each sex, where selection depends on the sex of the diploid they came from and their allele at the \textbf{A} locus.
Gametes/gametophytes from males then mate randomly with gametes/gametophytes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci.
Diploids then compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
This is followed by meiosis with recombination. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

%what were going to do, briefly
With allele $M$ fixed at the \textbf{M} locus, sex is determined by locus \textbf{X} and an equilibrium is reached at locus \textbf{A}. %where, importantly, locus \textbf{A} can be stably polymorphic. %the dynamics can be described by the frequency of $A$ in gametes/gametophytes from the homogametic sex and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex.
We then examine the ability of a rare, novel sex-determiner $m$ to increase in frequency at this equilibrium.
Numerical simulations are used to examine when an invading mutation goes to fixation.

@@

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p_{Xf}$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p_{Xm}$, and Y-bearing, $p_{Ym}$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
With no haploid selection and $k=0$ (neo-$Y$ invading a $ZW$ system) or $k=1$ (neo-$W$ invading an $XY$ system), our results reduce to those of \cite{vanDoorn:2010hu}.

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming equal dominance coefficients in the two sexes, no competition among gametes/gametophytes from females, and linear arrangement \textbf{X}\textbf{A}\textbf{M}.
Equivalent results for invasions into a $ZW$ system are given in the Sup.\ Mat.

\subsubsection*{Neo-$Y$}

The growth rate of a rare, dominant neo-$Y$ region ($k=0$) is
\begin{equation}
\lambda_{Y,XY} = 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
The neo-Y can therefore invade whenever it is in tighter linkage with the selected locus than is the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%write in general form without weak selection

\subsubsection*{Neo-$W$}

The growth rate of a rare, dominant neo-$W$ region ($k=1$) is
\begin{equation}
\lambda_{W,XY} = 1 + V_A \frac{s_f \left( (2 r (1-R) +R) s_f + (1- 2r) R s_m \right)}{2 r R \left( s_f + s_m \right)^2} t_m^2.
\end{equation}
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than is the ancestral sex-determining locus, $r<R$, a novel sex-determiner can invade.
For example, with $R=1/2$ the neo-W invades if there is some linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t_m\neq0$), and there is selection for or against $A$ in both males and females ($s_m s_f >0$) that is stronger in females than in males ($|s_f| > |s_m|$).
%write in general form without weak selection

\subsubsection*{Neo-ESD}

The growth rate of a rare, dominant neo-ESD region that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD,XY} = 1 + \frac{(\lambda_{Y,XY} - 1) + (\lambda_{W,XY} -1)}{4} \Big|_{R=1/2}
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s_m - 3 s_f\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\end{equation}
Thus the $k=1/2$ case gets half of the advantages of a neo-$W$ and half that of a neo-$Y$.  
Recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
%write in general form without weak selection

\linenumbers
\modulolinenumbers[2]
